---
title: "ARIMA_Module"
author: "Gabriel Vicencio, Max Dippel, Diego Larre and Miguel Garcia"
date: "11/29/2021"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading libraries
```{r}
library(nlme) # gls (generalized least squares)
library(MuMIn) #AICc
library(MASS) # acf(), pacf(), ARIMA()
library(ggplot2) # for graphing
library(curl) # for loading the data
library(forecast) # for predicting future values

```

### loading the dataset
```{r}
f <- curl("https://raw.githubusercontent.com/dalarre/Time-Series-Analysis/main/Dataset_R_2.csv")
d <- read.csv(f, header = TRUE, sep = ",", stringsAsFactors = FALSE)
head(d)
```
```{r}
names(d)
```

### graphing the dataset
```{r}
plot(d$Rb ~ d$X.U.FEFF.Year.AD, data=d, type="l", pch=16, 
     ylab="ppm", xlab="Year")
ggplot(data = d, aes(x=d$X.U.FEFF.Year.AD, y= d$Rb)) + geom_line()
```

### Analysis of the autocorrelation in the data
#### This is the pacf() function. This stands for partial auto-correlation function which is used to find auto-correlation in the data
```{r}


?pacf()
pacf1 <- pacf(d$Rb, main="PACF")
```

#### As you can see, there is auto correlation present in the data. It is shown by the data points which go beyond the blue confidence interval. Specifically the auto-correlation is outside the confidence interval at a lag of 1 and 2. This will guide later analysis. 

### Creating a simple linear model of the data
#### The next step in the analysis is to make a simple linear model of the data. We know this model does not accurately capture the variation in the data. However, it is helpful to look at the error present in this model to confirm that auto-correlation is present. 
```{r}
lm1 <- lm(d$Rb ~ d$X.U.FEFF.Year.AD, data=d)
summary (lm1)
```

### Analyzing autocorrelation in the residuals 
#### After that, we will extract the residuals (error) from this model and see if this error is auto-correlated as well
```{r}


d$mod.resid <- residuals(lm1)
pacf2 <- pacf(d$mod.resid, main="PACF - Rb residuals")
pacf2

```

#### Residuals should be random. The values in this graph which are outside the confidence interval tell us that the residuals are not random. The goal of this analysis will be to correct for these non-random residuals.


### Creating ARIMA Models
#### ARIMA stands for autoregressive intergrated moving average models. The values p, d and q are different values which help describe the data we are trying to fit to the ARIMA model. 
```{r}
### The arima() function in r is the model we will use to fit the data based on the parameters predicted by the pacf(). Specifically, since the values outside the confidence interval were a lag term of 1 and 2 we will use auto-regressive terms and moving average terms near 1-2. In general, it is good to expand a little beyond what is beyond the confidence interval because pacf() is not perfect. 

#AR
ar.1 <- arima(d$mod.resid, order=c(1,0,0)) 
ar.2 <- arima(d$mod.resid, order=c(2,0,0))
ar.3 <- arima(d$mod.resid, order=c(3,0,0))
ar.4 <- arima(d$mod.resid, order=c(4,0,0))

# MA
ma.1 <- arima(d$mod.resid, order=c(0,0,1)) 
ma.2 <- arima(d$mod.resid, order=c(0,0,2)) 
ma.3 <- arima(d$mod.resid, order=c(0,0,3)) 
ma.4 <- arima(d$mod.resid, order=c(0,0,4)) 

```

### Challenge 1: rank the ARIMA models using AICc
#### After fitting multiple ARIMAs we will use AICc() to rank the models from best to worst. The lower the AICc value the better it is 
```{r}


AICc(ar.1)
AICc(ar.2)
AICc(ar.3)
AICc(ar.4)
AICc(ma.1)
AICc(ma.2)
AICc(ma.3)
AICc(ma.4)
```
#### After running the AICc() we found that the models with 2 autoregressive terms and 3 moving average terms are the best fitted to the data. 

### Creating linear models which account for non-random error
#### We will now use the gls() function to model these two best fitting parameters into a linear model. gls stands for generalized least squares model. This type of linear model allows us to incorporate ARIMA terms into the modeling. 
```{r}


gls.mod1 <- gls(Rb ~ X.U.FEFF.Year.AD, data=d, correlation = corARMA(p=2, q=0))
gls.mod2 <- gls(Rb ~ X.U.FEFF.Year.AD, data=d, correlation = corARMA(p=0, q=3))
```

### Challenge 2: rank the models using AICc
```{r}
AICc(gls.mod1)
AICc(gls.mod2)
```
#### While the AICc value for model 1 is less than the model 2, they are so close we can consider than equivanlent. This means the data is either best modeled with 2 autoregressive terms or 3 moving average terms.  


### Checking our work  
#### Let's make a summary for the first linear model and see if the residuals still have auto-correlation using pacf()
```{r}
summary(gls.mod1)
confint(gls.mod1)

d$new.residuals <- resid(gls.mod1, type="normalized")

pacf3 <- pacf(d$new.residuals, main="PACF - GLS model residuals")
```

####  Poof! The auto-correlation is gone! Notice how there is now no values outside the confidence intervals. That means the data has been modeled in such a way that there is no autocorrelation error that we are not accounting for in the way we model the data. 

To make sure that the model is appropriate, we check if the residuals are normally distributed using a qq-plot

```{r}
qqnorm(d$new.residuals)
qqline(d$new.residuals)
```
The residuals adjust pretty well to the line in the Q-Q plot, which means they are normally distributed.

Next, we are going to use the model we created to make predictions of future values. We need the package "forecast" for this:

```{r}
forecast(ar.2,10)
```




